Efficiency

### 2018-09-09 Second.Rnw
some more

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
(Interview transcribed from Reference \parencite{lane2017}.)
Journalist: Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?
Jean-Michel Billig: It could be viable...
Journalist: So it's not mandated, its...
Jean-Michel Billig: What is driving the choice of technology is basically the operational requirement...
Journalist: So we could end up with conventional propeller?
Jean-Michel Billig: Yes, yes, we could, we could, we could also end up with something different
\\
\\
``However decisions on forms of propulsion are ultimately a matter for the Australian Government."\parencite{iggulden2017}
\end{tcolorbox}

@@

That DSTG is still working up from first principles to develop the capability to do ``independent analysis on the technologies" that we have accepted in a $50 billion dollar bid from a supposedly trusted design partner should come to all observers as a shocking revelation.  It is a deeply humiliating position for Australia, as a customer and an ally, to be so denied access to sufficient information to clearly and confidently understand the suitability of this technology.

@@

These series of comments by RADM Sammut are remarkeably consistent with other comments of his own and government sources, and wholly inconsitent with a competent and sound engineering and design process.

@@

One possibility is that a firm decision actually already has been reached on the inclusion or exclusion of these technologies, and RADM Sammut is not willing to reveal it.  It should be noted here that the 'current thinking' wording might be interpreted as something of a retreat from earlier 'we are using' statements made in late October 2017 \cite{sammut2017}.  The possibility that a pumpjet has already actually been ruled out must at least be considered.  In this scenario, it is almost certainly the case that the public and Parliament were misled on multiple occasions.

@@

However, such a possibility throws into sharp relieve some tensions between statements made by the government and Naval Group on multiple occasions.  RADM Sammut had previously stated that ``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements." \cite{sammut2017}.  Weeks earlier, Jean-Michel Billig had cited said ``What is driving the choice of technology is basically the operational requirement" when challenged as to whether the pumpjet was actually mandated \cite{lane2017}.  With both parties pleading ignorance of what technologies would ultimately be included, and both citing 'requirements', it does seem to imply that the requirements had not yet been settled.

If it was the case that the requirements had not been determined, then is likely that the Competitive Evaluation Process was entirely fraudulent.  If performance requirements weren't actually given to the competitors in order that their proposed solutions to meet them could be fairly and properly assessed, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}, and the credibility or conscience of the advisory panel who claimed they were should be seriously re-examined.  If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose.  A vast number of public statements made by politicians and officials regarding the significance of Australia's unique 'requirements', particularly for long range, should be critically re-examined.  It is likely that numerous statements made citing these requirements, and implying that they existed, were known, and were communicated to other parties including throughout and subsequent to the CEP, would be found to be misleading of the public, and the Parliament.

>>>>

\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.  Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
\\
\\
(Interview transcribed from Reference \parencite{lane2017}.)
Journalist: Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?

\textbf{Jean-Michel Billig:} It could be viable...

\textbf{Journalist:} So it's not mandated, its...

\textbf{Jean-Michel Billig:} What is driving the choice of technology is basically the operational requirement...

\textbf{Journalist:} So we could end up with conventional propeller?

\textbf{Jean-Michel Billig:} Yes, yes, we could, we could, we could also end up with something different
\\
\\
``However decisions on forms of propulsion are ultimately a matter for the Australian Government."\parencite{iggulden2017}
\end{tcolorbox}

@@

That DSTG is still working up from first principles to develop the capability to do ``independent analysis on the technologies" that we have accepted in a \$50 billion dollar bid from a supposedly trusted design partner should come to all observers as a shocking revelation.  It is a deeply humiliating position for Australia, as a customer and an ally, to be so denied access to sufficient information to clearly and confidently understand the suitability of this technology.

@@

These series of comments by RADM Sammut are consistent with other comments of his own and other government sources, and wholly inconsitent with a competent and sound engineering and design process.

@@

One possibility is that a firm decision actually already has been reached on the inclusion or exclusion of these technologies, and RADM Sammut is not willing to reveal it.  It should be noted here that the 'current thinking' wording might be interpreted as something of a retreat from earlier 'we are using' statements made in late October 2017 \parencite{sammut2017}.  The possibility that a pumpjet has already actually been ruled out must at least be considered.  In this scenario, it is almost certainly the case that the public and Parliament were misled on multiple occasions.

@@

However, such a possibility throws into sharp relieve some tensions between statements made by the government and Naval Group on multiple occasions.  RADM Sammut had previously stated that ``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements." \parencite{sammut2017}.  Weeks earlier, Jean-Michel Billig had cited said ``What is driving the choice of technology is basically the operational requirement" when challenged as to whether the pumpjet was actually mandated \parencite{lane2017}.  With both parties pleading ignorance of what technologies would ultimately be included, and both citing 'requirements', it does seem to imply that the requirements had not yet been settled.

If it was the case that the requirements had not been determined, then is likely that the Competitive Evaluation Process was entirely fraudulent.  If all the competitors weren't actually given the same performance requirements, which were the real requirements for the submarine which the government intended to procure, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}. Moreover, credibility or conscience of the advisory panel who claimed they were should be seriously re-examined.  If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose.  A vast number of public statements made by politicians and officials regarding the significance of Australia's unique 'requirements', particularly for long range, should be critically re-examined.  It is likely that numerous statements made citing these requirements, and implying that they existed, were known, and were communicated to other parties including throughout and subsequent to the CEP, would be found to be misleading of the public, and the Parliament.

The explanation given regarding AIP adds weight to the bizarre notion that a back-to-front design process is indeed taking place.  The enormous transit distances cited have been well known for decades, and presumeably the requirement to meet these had been distilled into specific ranges and endurances long before the commencement of the CEP.  It's true that large amounts of fuel and food might consequently be required.  But the idea that we 'might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumeably propososed by all contenders alongside an AIP system were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further.  It would be relatively simple at the very early stages of concept design (prior to the diameter and legnth of the submarine being fixed) to incorporate both sufficient fuel and food for a specific overall range and endurance requirement, as well as an AIP system to enhance the dived endurance, without compromising the overall range and endurance.  That this was not done, and instead the dimensions of the hull were fixed, whilst Navy continued to search for a system which might fit within this pre-defined platform envelope does not remotely resemble a competent strategy to acquire a capability to meet a fixed requirement.

Worryingly, it adds confirmation that RADM Sammut might genuinely believe that the hull-diameter (and its resemblance of a pre-existing design) ought to be a pre-determined feature of a submarine design, to which the performance and capability of the submarine is subordinate.  Such a view, including held on the grounds of safety as suggested by RADM Sammut himself \ref{SENhanson}, is such a bizarre understanding of program risk and the heirachy of engineering challenges that it should call into question the competence of a program manager, and the technical advisors who he depends upon.

The competence of RADM Sammut's program management and the technical advisors on whom he depends is further marred by the comment about lithium-ion batteries being now firmly excluded from the design because the ``RAN does not want to be introducing a massive number of new technologies simultaneously".  This is telling because lead-acid batteries are known to have a relatively high internal resistance, which increases as they become hot, which occurs during levels of faster discharge.  This means that the efficiency of the batteries falls substantially at higher rates of discharge, reducing substantially the total amount of stored energy that they can usefully recover at very high discharge rates.  Whilst this is generally true for all batteries, Lithium-ion batteries are known to have substantially lower internal resistance, and consequently have dramatically improved performance (in terms efficiency) at higher rates of discharge when  compared to lead-acid batteries.  It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid at high speeds \parencite{briggs2016}.  Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.  Without them, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.  This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to under one hour.  With lithium ion batteries, it is plausible that slightly over 2 hours at this speed could be sustained.

As I have also pointed out
